\textit{k}-mer-based GWAS enhances the discovery of causal variants and candidate genes in soybean

### 2023-07-05 additional_files/manuscript.tex
Add a figure about sequencing depth and some material in the discussion.

We used the approach outlined by \shortciteA{voichek2020} for \emph{k}-mer-based GWAS
and their recommendations on the
associated GitHub page (v. 0.2-beta, \url{https://github.com/voichek/kmersGWAS}) to generate
a presence/absence \emph{k}-mer table for use in GWAS.  The first step of
the approach involved counting all \emph{k}-mers of length 31 (31-mers) present
in the trimmed (FASTQ) reads for each accession using KMC3 v. 3.2.1
\shortcite{kokot2017}. We selected $k = 31$ based on demonstrated performance in previously
published studies \shortcite{rahman2018, voichek2020, he2021}.
\emph{k}-mers were counted twice, the first time using their
canonized form (the first of either the observed \textit{k}-mer or its reverse
complement in lexicographical order) and the second time using the observed
\emph{k}-mers themselves (non-canonized form). The two sets of \textit{k}-mers
were then combined and filtered by
keeping only those seen in at least 5 accessions, and observed in both canonized
and non-canonized form in at least 20\% of the accessions in which they are found.
We then generated a table indicating the presence or absence of each
\emph{k}-mer in all accessions. This table was used to compute a kinship
table for use in GWAS and was also used directly as input genotypes for the
GWAS analyses. We used all \emph{k}-mers with a MAF $\geq$ 0.02 as input genotypes
for both the computation of kinship and GWAS analysis. The total number of \textit{k}-mers
in this filtered table was 637,605,529.

@@

While we have have shown that using \emph{k}-mers for GWAS in soybean performed
well compared to other GWAS methods, this approach does present some unique
challenges and limitations. One obvious limitation is that using the
presence/absence of \emph{k}-mers may not be appropriate in species where
heterozygosity is common (i.e., most species other than inbred crops) as the
\emph{k}-mers associated with the presence of both alleles will be observed at
heterozygous loci. Moreover, although the presence/absence of \emph{k}-mers can
detect variation independently of its type, some types of variation may be more
difficult to detect than others.  For example, copy number variation at a locus
may not be detected from presence/absence alone if all copies are identical and
occur in tandem, such that no \textit{k}-mer can differentiate between two
copies and more. For similar reasons, \emph{k}-mers may also fail to capture
variation occurring in repeated sequences. Alternative models based on
\emph{k}-mer counts instead of presence/absence have been developed
\shortcite{rahman2018, he2021} which may solve issues like these and broaden the
scope of variants that can be identified from \emph{k}-mers.
Nevertheless, \shortciteA{voichek2020} demonstrated that 98.9\% of known translocations and
inversions as well as 99.6\% of known insertions/deletions in \textit{Arabidopsis thaliana}
could be tagged by unique 31-mers. It is therefore likely that the vast majority
of variants can be detected through the presence/absence of \textit{k}-mers.

>>>>

We used the approach outlined by \shortciteA{voichek2020} for \emph{k}-mer-based GWAS
and their recommendations on the
associated GitHub page (v. 0.2-beta, \url{https://github.com/voichek/kmersGWAS}) to generate
a presence/absence \emph{k}-mer table for use in GWAS.  The first step of
the approach involved counting all \emph{k}-mers of length 31 (31-mers) present
in the trimmed (FASTQ) reads for each accession using KMC3 v. 3.2.1
\shortcite{kokot2017}. We selected $k = 31$ based on demonstrated performance in previously
published studies \shortcite{rahman2018, voichek2020, he2021}.
\emph{k}-mers were counted twice, the first time using their
canonized form (the first of either the observed \textit{k}-mer or its reverse
complement in lexicographical order) and the second time using the observed
\emph{k}-mers themselves (non-canonized form). The two sets of \textit{k}-mers
were then combined and filtered by
keeping only those seen in at least 5 accessions, and observed in both canonized
and non-canonized form in at least 20\% of the accessions in which they are found.
Only \textit{k}-mers that occurred twice in a given sample were considered to
reduce the impact of sequencing errors.
We then generated a table indicating the presence or absence of each
\emph{k}-mer in all accessions. This table was used to compute a kinship
table for use in GWAS and was also used directly as input genotypes for the
GWAS analyses. We used all \emph{k}-mers with a MAF $\geq$ 0.02 as input genotypes
for both the computation of kinship and GWAS analysis. The total number of \textit{k}-mers
in this filtered table was 637,605,529.

@@

While we have have shown that using \emph{k}-mers for GWAS in soybean performed
well compared to other GWAS methods, this approach does present some unique
challenges and limitations. One of these challenges is the
interpretation of \textit{k}-mer-based GWAS given the greater possibility
of false positive associations than in conventional GWAS approaches. In this
study, we noticed apparently spurious associations with \textit{k}-mers
scattered throughout the genome were common. This was especially the case for
highly significant loci whose signal spanned several megabases such as \textit{W}
and \textit{T}. A look at the Q-Q plots derived from \textit{k}-mer-based GWAS
showed that these usually had the same shape as the Q-Q plots derived from SNP-
and SV-based GWAS, suggesting that population structure was properly controlled.
We believe that a more likely culprit for these spurious associations is the
difficulty in linking \textit{k}-mers to genomic coordinates. At loci that
span a large interval, it is indeed be highly likely that some of the \textit{k}-mers
coming from the region map elsewhere in the genome. This interpretation is supported
by the fact that the LD between presumably spurious \textit{k}-mers and
\textit{k}-mers from target loci was generally high (e.g. Figures S5 and S47).
Overall, the risk of false association due to population structure and the identification
of an appropriate significance threshold seems well controlled, but work remains
to be done in reducing the occurrence of spurious associations.

There are obvious limitations to the relying on the presence/absence of
\emph{k}-mers for GWAS. For example, this approach may not be appropriate in
species where heterozygosity is common (i.e., most species other than inbred
crops) as the \emph{k}-mers associated with the presence of both alleles will
be observed at heterozygous loci. Moreover, although the presence/absence of
\emph{k}-mers can detect variation independently of its type, some types of
variation may be more difficult to detect than others.  For example, copy
number variation at a locus may not be detected from presence/absence alone if
all copies are identical and occur in tandem, such that no \textit{k}-mer can
differentiate between two copies and more. For similar reasons, \emph{k}-mers
may also fail to capture variation occurring in repeated sequences. Alternative
models based on \emph{k}-mer counts instead of presence/absence have been
developed \shortcite{rahman2018, he2021} which may solve issues like these and
broaden the scope of variants that can be identified from \emph{k}-mers.
Nevertheless, \shortciteA{voichek2020} demonstrated that 98.9\% of known
translocations and inversions as well as 99.6\% of known insertions/deletions
in \textit{Arabidopsis thaliana} could be tagged by unique 31-mers. It is
therefore likely that the vast majority of variants can be detected through the
presence/absence of \textit{k}-mers.

In addition to the link between \textit{k}-mer presence/absence and sequence
variation, one may wonder to what extent errors may occur in declaring the
presence or absence of a \textit{k}-mer based on Illumina WGS data. If we
neglect the possibility of contamination, the most likely error would be to
declare a \textit{k}-mer as absent simply because sequencing depth was
insufficient. In this study, we declared \textit{k}-mers as present
based on a minimum of two occurrences. Therefore, a \textit{k}-mer may be
wrongly declared as absent if it is observed only once or not at all.  To
quantify this error, we compared the number of times that a given position was
observed only once in a sample to the average sequencing depth of that sample.
We found that most samples with a sequencing depth between 10 and 20X had
a proportion of only 0.5 to 1.5\% of positions covered by only one read.
Therefore, the risk of falsely declaring a \textit{k}-mer as absent from
our dataset was low, and in any case much lower than typical error rates
involved in calling and genoyping SVs \shortcite<e.g.>[]{kosugi2019, ebler2022}.
The robustness of \textit{k}-mer presence/absence to sequencing depth is also
demonstrated by \shortciteA{gaurav2022}.
